---
title: "ISLR Chapter 2 Exercises"
date: 2023-07-01
output: html_document
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Table of Contents
1. [Question 1](#Question-1)
2. [Question 2](#Question-2)
3. [Question 3](#Question-3)
4. [Question 4](#Question-4)
5. [Question 5](#Question-5)
6. [Question 6](#Question-6)
7. [Question 7](#Question-7)
8. [Question 8](#Question-8)
9. [Question 9](#Question-9)
10. [Question 10](#Question-10)


## Conceptual

### Question 1

#### a.
We would expect a flexible statistical learning method to perform better than
an inflexible method because the risk of overfitting is minimal with a large sample
size and small number of predictors. A flexible method will thus have lower bias
and negligibly higher variance than an inflexible method.

### b.
We would expect an inflexible statistical learning method to perform better than a
flexible method because the flexible method will be at risk of overfitting with a small sample size
and a large number of predictors. The inflexible method will have higher bias but much lower variance
than the flexible method in this case.

### c.
We would expect the flexible method to perform better here because it will be able to
learn the non-linear relationship between the predictors and the response variable better than the inflexible method.
The flexible method will thus have much lower bias than the inflexible method, offsetting the increase in variance.


### d.
We would expect the inflexible method to perform better here because the flexible method
will likely model the large error terms rather than the underlying true relationship between the response
variable and the predictors. The flexible method will have very high variance in this case.

### Question 2

#### a.
This is a regression problem, as CEO salary is a continuous variable. We are most
interested in inference here. `n` = 500, `p` = 3.

#### b.
This is a classification problem, as the response variable is binary. We are most interested in prediction.
`n` = 20, `p` = 13.

#### c.
This is a regression problem, as %change is a continuous variable. We are most interested in prediction.
`n` = 52 (52 weeks in a year), `p` = 3.

### Question 3

```{r error_source_graph, }

df_bias_squared <- data.frame(
  flexibility = c(2, 15, 25),
  error = c(1, 0.008, 0.005)
)
df_bias_squared$source <- "bias_squared"

df_variance <- data.frame(
  flexibility = c(2, 15, 25),
  error = c(0.01, 0.25, 1)
)
df_variance$source <- "variance"

df_training_error <- data.frame(
  flexibility = c(2, 10, 25),
  error = df_bias_squared$error * 0.8
)
df_training_error$source <- "training_error"

df_irreducible_error <- data.frame(
  flexibility = c(0, 25),
  error = c(0.5, 0.5)
)
df_irreducible_error$source <- "irreducible_error"

df_total_error <- data.frame(
  flexibility = c(2, 15, 25),
  error = df_bias_squared$error + df_variance$error + df_irreducible_error$error[[1]]
)
df_total_error$source <- "test_error"

df_errors <- dplyr::bind_rows(df_bias_squared, df_variance, df_training_error, df_total_error)

ggplot2::ggplot(df_errors) +
  ggplot2::geom_smooth(ggplot2::aes(x = flexibility, y = error, color = source)) +
  ggplot2::geom_line(data = df_irreducible_error, ggplot2::aes(x = flexibility, y = error, color = source))
```

## Applied

Load libraries.
```{r load_libraries}
library(magrittr)
library(dplyr)
library(GGally)
library(ggplot2)
library(tidyr)
library(MASS)
library(ISLR)
```

Summarize data.
```{r summarize_data}
df_college <- College

summary(df_college)

```
