---
title: "ISLR Chapter 2 Exercises"
date: 2023-07-01
output: html_document
---

# Table of Contents
1. [Question 1](#Question_1)
2. [Question 2](#Question_2)
3. [Question 3](#Question_3)
4. [Question 4](#Question_4)
5. [Question 5](#Question_5)
6. [Question 6](#Question_6)
7. [Question 7](#Question_7)
8. [Question 8](#Question_8)
9. [Question 9](#Question_9)
10. [Question 10](#Question_10)


## Conceptual

### Question_1

#### a.
We would expect a flexible statistical learning method to perform better than
an inflexible method because the risk of overfitting is minimal with a large sample
size and small number of predictors. A flexible method will thus have lower bias
and negligibly higher variance than an inflexible method.

### b.
We would expect an inflexible statistical learning method to perform better than a
flexible method because the flexible method will be at risk of overfitting with a small sample size
and a large number of predictors. The inflexible method will have higher bias but much lower variance
than the flexible method in this case.

### c.
We would expect the flexible method


## Applied

Load libraries.
```{r load_libraries}
library(magrittr)
library(dplyr)
library(GGally)
library(ggplot2)
library(tidyr)
library(MASS)
library(ISLR)
```

Summarize data.
```{r summarize_data}
df_college <- College

summary(df_college)

```
